---
title: "Human Activity Recognition"
author: "Baldvin Einarsson"
date: "6/3/2017"
output: 
    html_document:
        keep_md: TRUE

---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment="")
library(knitr)
```

Human Activity Recognition (HAR) is an interesting topic, with monitors and sensors becoming widely used. See information on original data in the following website:
[http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har)

Here, we look at the weight lifting exercise dataset. This dataset is concerned with classifying whether an a weight lifting exercise was done properly.

# Data

Let's download training and test sets from the following websites:

* The training data:
[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

* The test data:
[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

Let's import these data files into `R`, into objects `training` and `test`.
```{r loadData, echo=FALSE}
training <- read.csv(file = "./data/pml-training.csv",
                     header = TRUE, stringsAsFactors = FALSE)

test <- read.csv(file = "./data/pml-testing.csv",
                     header = TRUE, stringsAsFactors = FALSE)
```

The data contains `r length(training)` columns. Six people performed 10 repetitions of the "Unilateral" Dubbell Biceps Curl", in five different ways. The variable we're interested is called `classe`, and the unique values labeled as A through E, with the following descriptions:

The classes stand for the following:
```{r describeClasses,echo=FALSE}
kable(data.frame(Class=LETTERS[1:5],
                 Description = c("Exactly according to the specification",
                                 "Throwing the elbows to the front",
                                 "Lifting the dumbbell only halfway",
                                 "Lowering the dumbbell only halfway",
                                 "Throwing the hips to the front")
                 )
      )
```

